---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# selection.index

<!-- badges: start -->
[![R-CMD-check](https://github.com/zankrut20/selection.index/workflows/R-CMD-check/badge.svg)](https://github.com/zankrut20/selection.index/actions)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/zankrut20/selection.index?branch=master&svg=true)](https://ci.appveyor.com/project/zankrut20/selection.index)
[![CRAN checks](https://cranchecks.info/badges/summary/selection.index)](https://cran.r-project.org/web/checks/check_results_selection.index.html)
[![CRAN status](https://www.r-pkg.org/badges/version/selection.index)](https://CRAN.R-project.org/package=selection.index)
[![Dependencies](https://tinyverse.netlify.com/badge/selection.index)](https://cran.r-project.org/package=selection.index)
[![Total downloads](http://cranlogs.r-pkg.org/badges/grand-total/selection.index?color=blue)](https://cran.r-project.org/package=selection.index)
[![Last-Month downloads](http://cranlogs.r-pkg.org/badges/last-month/selection.index?color=green)](https://cran.r-project.org/package=selection.index)
[![Last-Week downloads](http://cranlogs.r-pkg.org/badges/last-week/selection.index?color=green)](https://cran.r-project.org/package=selection.index)
[![CodeFactor](https://www.codefactor.io/repository/github/zankrut20/selection.index/badge)](https://www.codefactor.io/repository/github/zankrut20/selection.index)
<!-- badges: end -->

The goal of selection.index is to easily construct the selection index and based on the these indices select the plant traits for the overall improvement of the plant.

## Installation

You can install the released version of selection.index from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("selection.index")
```

## Example

This is a basic example which shows you how to solve a common problem:
Dataset selindexdata is included in package. 

```{r example}
library(selection.index)
head(seldata)
```
### Genotypic Variance-Covariance Matrix
```{r}
genMat<- gen.varcov(data = seldata[,3:9], genotypes = seldata[,2],
                    replication = seldata[,1])
print(genMat)
```

### Phenotypic Variance-Covariance Matrix
```{r}
phenMat<- phen.varcov(data = seldata[,3:9], genotypes = seldata[,2],
                      replication = seldata[,1])
print(phenMat)
```

+ Genetic gain of Yield
```{r}
GAY<- gen.advance(phen_mat = pmat[1,1], gen_mat = gmat[1,1],
                  weight_mat = w[1,2])
print(GAY)
```

### Construction of selection index/indices
For the construction of selection index we requires **phenotypic & genotypic variance-covariance matrix as well weight matrix.**<br>

### Construction of all possible selection indices for a character combinations
```{r}
comb.indices(ncomb = 1, pmat = phenMat, gmat = genMat, wmat = weight[,2:3], wcol = 1, GAY = GAY)
```

### Construction of selection indices by removing desired character from the combinations
```{r}
rcomb.indices(ncomb = 1, i = 1, pmat = phenMat, gmat = genMat, wmat = weight[,2:3], wcol = 1, GAY = GAY)
```

