---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# selection.index

<!-- badges: start -->
[![CRAN checks](https://cranchecks.info/badges/summary/selection.index)](https://cran.r-project.org/web/checks/check_results_selection.index.html)
[![R-CMD-check](https://github.com/zankrut20/selection.index/workflows/R-CMD-check/badge.svg)](https://github.com/zankrut20/selection.index/actions)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/zankrut20/selection.index?branch=master&svg=true)](https://ci.appveyor.com/project/zankrut20/selection.index)
[![CRAN status](https://www.r-pkg.org/badges/version/selection.index)](https://CRAN.R-project.org/package=selection.index)
[![Dependencies](https://tinyverse.netlify.com/badge/selection.index)](https://cran.r-project.org/package=selection.index)
[![CRAN downloads](http://cranlogs.r-pkg.org/badges/last-month/selection.index?color=green)](https://cran.r-project.org/package=selection.index)
<!-- badges: end -->

The goal of selection.index is to easily construct the selection index and based on the these indices select the plant traits for the overall improvement of the plant.

## Installation

You can install the released version of selection.index from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("selection.index")
```

## Example

This is a basic example which shows you how to solve a common problem:
Dataset selindexdata is included in package. 

```{r example}
library(selection.index)
head(seldata)
```
### Genotypic Variance-Covariance Matrix
```{r}
genMat<- gen.varcov(data = seldata[,3:9], genotypes = seldata[,2],
                    replication = seldata[,1])
print(genMat)
```

### Phenotypic Variance-Covariance Matrix
```{r}
phenMat<- phen.varcov(data = seldata[,3:9], genotypes = seldata[,2],
                      replication = seldata[,1])
print(phenMat)
```

### Construction of selection index/indices
For the construction of selection index we requires **phenotypic & genotypic variance-covariance matrix as well weight matrix.**<br>

+ In **"selection.index"** function **"GAY"** is optional imput. The same function is used to calculate the genetic gain of desired trait/character's selection index. 

```{r}
s<- list()
s[[1]]<- sel.index(ID = 1, phen_mat = phenMat[1,1], 
                   gen_mat = genMat[1,1], weight_mat = weight[1,2])
```

+ Based on above GAY value we calculate the further selection index for the other traits/characters.
```{r}
s[[2]]<- sel.index(ID = 2, phen_mat = phenMat[2,2],
                   gen_mat = genMat[2,2], weight_mat = weight[2,2], 
                   GAY = 2.1468)
```

### Ranking of selection indices
```{r}
r<- rank.index(s,2)
print(r)
```

### Selection score and Ranking of genotypes
```{r}
sr<- sel.score.rank(data = seldata[,3], bmat = 0.5, genotype = seldata[,2])
head(sr)
```
