---
title: "Data Analysis with selection.index"
author: "Zankrut Goyani"
date: "`r Sys.Date()`"
time: "`r Sys.time()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Analysis with selection.index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The aim of most plant breeding program is simultaneous improvement of several characters. An objective method involving simultaneous selection for several attributes then becomes necessary. It has been recognized that most rapid improvements in the economic value is expected from selection applied simultaneously to all the characters which determine the economic value of a plant, and appropriate assigned weights to each character according to their economic importance, heritability and correlations between characters. So the selection for economic value is a complex matter. If the component characters are combined together into an index in such a way that when selection is applied to the index, as if index is the character to be improved, most rapid improvement of economic value is excepted. Such an index was first proposed by Smith (1937) based on the Fisher's (1936) "discriminant function". In this package selection index is calculated based on the Smith (1937) selection index method (Dabholkar, 1999). For more information refer **Elements of Bio Metrical GENETICS by A. R. Dabholkar.**
```{r setup}
library(selection.index)
d<- seldata # Manually generated data for analysis which is included in package
```

```{r}
w<- weight # Weights assigned to the traits also include in package
```

As we discussed that selection index based on discriminant function. So we have required **genotypic & phenotypic variance-covariance matrix** for further analysis.

+ Genotypic variance-covariance matrix
```{r}
gmat<- gen_varcov(data = d[,3:9], genotypes = d$treat, replication = d$rep)
print(gmat)
```

+ Phenotypic variance-covariance matrix
```{r}
pmat<- phen_varcov(data = d[,3:9], genotypes = d$treat, replication = d$rep)
print(pmat)
```

Generally, **Percent Relative Efficiency (PRE)** of a selection index is calculated with reference to **Genetic Advance (GA) yield** of respective weight. So first we calculate the GA of yield for respective weights.
+ Genetic gain of Yield
```{r}
GAY<- gen_advance(phen_mat = pmat[1,1], gen_mat = gmat[1,1],
                  weight_mat = w[1,2])
print(GAY)
```
We use this GAY value for the construction, ranking of the other selection indices and stored them in a list "si".

## Selection score and Ranking of genotypes
Generally selection score is calculate based on top ranked selection index. So first we store the **discriminant coefficient** value into a variable **b**, and later that value we used for calculation of selection score and ranking of the genotypes.

## `lpsi()` is used for construction of selection indices based on different combination of characters. 
```{r}
lpsi(ncomb = 1, pmat = pmat, gmat = gmat, wmat = w[,-1], wcol = 1, GAY = GAY)
```

## Using `excluding_trait` parameter to remove traits from the construction of selection indices
```{r}
lpsi(ncomb = 1, pmat = pmat, gmat = gmat, wmat = w[,-1], wcol = 1, GAY = GAY, excluding_trait = 1)
```

## Genomic Selection Workflows

Modern plant breeding increasingly relies on genomic prediction to estimate breeding values from molecular markers. The package provides specialized functions for computing genomic variance-covariance matrices used in genomic selection indices.

### Genomic Variance-Covariance Matrix (Γ)

The `genomic_varcov()` function computes the genomic variance-covariance matrix from Genomic Estimated Breeding Values (GEBVs):

```{r, eval=FALSE}
# Simulate GEBVs from genomic prediction
set.seed(123)
n_genotypes <- 100
n_traits <- 5
gebv_mat <- matrix(rnorm(n_genotypes * n_traits), 
                   nrow = n_genotypes, ncol = n_traits)
colnames(gebv_mat) <- paste0("Trait", 1:n_traits)

# Compute genomic variance-covariance matrix
Gamma <- genomic_varcov(gebv_mat)
print(Gamma)
```

**Handling Missing Values in GEBVs:**

When genomic predictions contain missing values, specify the `use` parameter:

```{r, eval=FALSE}
# Use complete observations only (recommended)
Gamma <- genomic_varcov(gebv_mat, use = "complete.obs")

# Pairwise deletion (may not be PSD - use with caution)
Gamma <- genomic_varcov(gebv_mat, use = "pairwise.complete.obs")
```

**Important:** Pairwise deletion can produce matrices that are not positive semi-definite (PSD), which may cause numerical issues in selection index calculations. Use `complete.obs` when possible.

### Phenomic-Genomic Covariance Matrix (Φ)

The `phenomic_genomic_varcov()` function computes the covariance between phenotypic observations and GEBVs:

```{r, eval=FALSE}
# Compute phenomic-genomic covariance
Phi <- phenomic_genomic_varcov(pheno_mat = d[,3:9], 
                                gebv_mat = gebv_mat)
print(Phi)
```

This matrix is used in selection indices that combine phenotypic and genomic information.

### Genetic-Genomic Covariance Matrix (A)

The `genetic_genomic_varcov()` function computes the covariance between genetic values (estimated from field trials) and GEBVs:

```{r, eval=FALSE}
# Compute genetic-genomic covariance
A_mat <- genetic_genomic_varcov(gv_mat = gmat,  # genetic values
                                 gebv_mat = gebv_mat)
print(A_mat)
```

**Applications in Genomic Selection Indices:**

- **LGSI (Linear Genomic Selection Index):** Uses Γ (genomic variance)
- **Combined Indices:** Use Φ (phenomic-genomic covariance) and A (genetic-genomic covariance)
- **Multi-environment:** Combine field trial covariances with genomic predictions

## Missing Value Imputation

Field trials often contain missing observations due to experimental errors, pest damage, or environmental stress. The `impute_missing()` function estimates missing values using the experimental design structure.

### Basic Usage

```{r, eval=FALSE}
# Create data with missing values
d_missing <- d
d_missing[c(5, 12, 20), 3] <- NA  # Add missing values to first trait

# Impute missing values
d_complete <- impute_missing(data = d_missing[,3:9], 
                              genotypes = d_missing$treat,
                              replications = d_missing$rep,
                              design = "RCBD",
                              method = "REML")

# Check: no missing values remain
any(is.na(d_complete))
```

### Method Selection Guide

Six imputation methods are available, with different strengths:

| Method | Speed | Accuracy | Missing Pattern | Design Support |
|--------|-------|----------|-----------------|----------------|
| **REML** | Slow | Highest | Complex | RCBD, LSD |
| **Yates** | Fast | High | Simple | RCBD, LSD |
| **Healy** | Medium | High | Multiple per block | RCBD, LSD |
| **Regression** | Fast | High | Any | RCBD, LSD |
| **Mean** | Fastest | Medium | Any | RCBD, LSD, SPD |
| **Bartlett** | Medium | Highest* | Any | RCBD, LSD |

*When traits are correlated

**Recommendations:**

- **REML:** Best for complex missing patterns or when precision is critical
- **Yates:** Simple, fast, good for few missing values in balanced designs
- **Healy:** More stable than Yates when multiple values missing
- **Regression:** Fast and deterministic, good general-purpose choice
- **Mean:** Quick estimation when precision less critical
- **Bartlett:** Use when traits are highly correlated (leverages trait relationships)

### Design-Specific Usage

**Latin Square Design:**
```{r, eval=FALSE}
d_complete <- impute_missing(data = d_missing[,3:9],
                              genotypes = d_missing$treat,
                              replications = d_missing$row,  # row indices
                              columns = d_missing$col,       # column indices
                              design = "LSD",
                              method = "Healy")
```

**Split Plot Design:**
```{r, eval=FALSE}
d_complete <- impute_missing(data = d_missing[,3:9],
                              genotypes = d_missing$subplot,    # sub-plot treatments
                              replications = d_missing$block,
                              main_plots = d_missing$mainplot,  # main plot treatments
                              design = "SPD",
                              method = "Mean")  # Only Mean available for SPD
```

### Workflow Integration

After imputation, proceed with variance-covariance analysis:

```{r, eval=FALSE}
# 1. Impute missing values
d_complete <- impute_missing(data = d[,3:9],
                              genotypes = d$treat,
                              replications = d$rep,
                              design = "RCBD",
                              method = "REML")

# 2. Compute variance-covariance matrices
gmat <- gen_varcov(data = d_complete, 
                   genotypes = d$treat, 
                   replication = d$rep)
                   
pmat <- phen_varcov(data = d_complete,
                    genotypes = d$treat,
                    replication = d$rep)

# 3. Construct selection indices
GAY <- gen_advance(phen_mat = pmat[1,1], gen_mat = gmat[1,1],
                   weight_mat = w[1,2])
                   
lpsi(ncomb = 1, pmat = pmat, gmat = gmat, wmat = w[,-1], 
     wcol = 1, GAY = GAY)
```

**Best Practices:**

1. **Examine missing patterns** before imputation - understanding why data are missing helps choose appropriate methods
2. **Use REML or Healy** for complex or extensive missingness
3. **Document imputation method** in analysis reports
4. **Consider multiple imputation** for high proportions of missing data (>20%)
5. **Adjust degrees of freedom** in final analysis to account for imputed values
